---
title: "Hands-on Exercise 5: Geographical Segmentation with Spatially Constrained Clustering Techniques"
editor: visual
---

## 1 Overview

In this exercise, you will gain hands-on experience on how to delineate homogeneous region by using geographically referenced multivariate data. There are two main analysis, namely:

-   hierarchical cluster analysis, and

-   spatially constrained cluster analysis

### 1.1 Learning Outcome

By the end of this hands-on exercise, the following can be achieved:

-   to convert GIS polygon data into R's simple feature data.frame by using appropriate functions of **sf** package of R;

-   to convert simple feature data.frame into R's SpatialPolygonDataFrame object by using appropriate **sf** of package of R;

-   to perform custer analysis by using *hclust()* of Base R;

-   to perform spatially constrained cluster analysis using *skater()* of Base R; and

-   to visualise the analysis output by using **ggplot2** and **tmap** package.

## 2 Getting Started

### 2.1 The analytical question

In geobusiness and spatial policy, it is a common practice to delineate the market or planning area into homogeneous regions by using multivariate data. In this hands-on exercise, we are interested to delineate [Shan State](https://en.wikipedia.org/wiki/Shan_State), [Myanmar](https://en.wikipedia.org/wiki/Myanmar) into homogeneous regions by using the penetration of multiple Information and Communication Technology (ICT) measures, namely: Radio, Television, Landl line ohone, Mobile phone, Computer, and Internet at home.

## 3 The Data

Two data sets will be used in this study. They are:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*): This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both datasets are downloaded from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

### 3.1 Installing and loading R packages

Before we get started, it is important for us to install the necessary R packages into R and launch these R packages into R environment. The R packages needed for this exercise are as follows:

-   Spatial data handling

    -   **sf**, **rgdal** and **spdep**

-   Attribute data handling

    -   **tidyverse** (which contains **readr**, **ggplot2** and **dplyr**)

-   Choropleth mapping

    -   **tmap**

-   Multivariate data visualisation and analysis

    -   **coorplot**, **ggpubr**, and **heatmaply**

-   Cluster analysis

    -   **cluster**

    -   **ClustGeo**

The code chunks below installs and launches these R packages into R environment.

```{r}
pacman::p_load(rgdal, spdep, tmap, sf, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse)
```

## 4 Data Import and Preparation

### 4.1 Importing geospatial data into R environment

In this section, we will import Myanmar Township Boundary GIS data and its associated attribute table into R environment.

The Myanmar Township Boundary GIS data is in ERSI shapefile format. It will be imported into R environment by using the [*st_read()*](https://www.rdocumentation.org/packages/sf/versions/0.7-2/topics/st_read) function of **sf**.

The code chunks used are shown below:

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)"))
```

The imported township boundary object is called *shan_sf*. It is saved in **simple feature data.frame** format. We can view the content of the newly created *shan_sf* simple features data.frame by using the code chunk below.

```{r}
shan_sf
```

Since sf.data.frame conforms to the tidy framework, we can use `glimpse`()` to reveal the data type in each field in *shan_sf*.

### 4.2 Importing aspatial data into R environment

The csv file will be imported using read_csv  frunction of *readr* package as shown in the following code chunk:

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from The 2014 Myanmar Population and Housing Census Myanmar. The attribute data set is called ict. It is saved in R’s tibble data.frame format.

The code chunk below shows the summary statistics of ict data.frame.

```{r}
summary(ict)
```

We can see that there a total of eleven fields and 55 observations in the tibble data.frame.

### 4.3 Deriving new variables uing **dplyr** package

The data in *ict* provides the count of the number of households. Such units of measurement is directly biased by the underlying total number of households in the town. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV etc. 

In order to overcome this issue, we will derive the penetration rate for each ICT by using the code chunk below. 

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,
         `DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, 
         `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, 
         `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, 
         `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, 
         `INTERNET`=`Internet at home`) 
```

We can review the summary statistics of the newly derived variables using the code chunk below. 

```{r}
summary(ict_derived)
```

## 5 Exploratory Data Analysis (EDA)

### 5.1 EDA using statistical graphs









